proof-of-concept_mods

### 2023-10-31 proof-of-concept_mods.Rmd
centile diffs by sex

```{r}
drop1.df <- fread("ukb_basic/drop1_tests.csv", stringsAsFactors = TRUE) %>%
  dplyr::select(-V1)
sex.df <- fread("ukb_basic/sex_summary.csv", stringsAsFactors = TRUE) %>%
  dplyr::select(-V1)
sum.df <- fread("ukb_basic/gamlss_summary.csv", stringsAsFactors = TRUE)%>%
  dplyr::select(-V1)
```

@@

sigma.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & term =="sexMale") %>%
  group_by(dataset, pheno_cat) %>%
  ggplot() +
  geom_histogram(aes(x=var_est, fill = dataset)) +
  facet_wrap(~pheno_cat, scales= "free")
```

@@

```{r, fig.height=6, fig.widt=8}
sigma.df %>%
  dplyr::filter(pheno_cat == "Global Volume" & term =="sexMale") %>%
  mutate(var_est.sig = case_when(sig.sum_bf.corr == TRUE ~ var_est,
                                 sig.sum_bf.corr == FALSE ~ 0)) %>%
  #group_by(dataset) %>%
  ggplot() +
  geom_col(aes(x=dataset, y = var_est.sig, fill=dataset), position="dodge") +
  #scale_fill_gradient2() +
  facet_grid(~pheno) + 
  theme_bw(base_size=16) +
  #theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_rect(color= "gray", fill = NA)) +
  ggtitle("Beta weight for effect of male sex on global volumes' variance")

sigma.df.glob <- sigma.df %>%
  dplyr::filter(pheno_cat == "Global Volume" & term =="sexMale")
table(sigma.df.glob$dataset, sigma.df.glob$pheno)

sigma.df %>%
  dplyr::filter(pheno == "sGMV")
```

@@

table(combined_preds$dataset)
names(combined_preds)
```
calc how much centiles change
```{r}
cent_diffs <- combined_preds %>%
  mutate(dataset = factor(dataset, levels = unique(dataset), ordered = FALSE)) %>%
  mutate(dataset = relevel(dataset, ref= "ukb_CN")) %>%
  arrange(dataset) %>%
  group_by(participant) %>%
  mutate(across(all_of(pheno_list), ~ . - first(.), .names = "diff_{.col}")) %>%
  ungroup()

diff_vol_glob <- sapply(vol_list_global, function(x) paste0("diff_", x))

mean.cent.diffs <- cent_diffs %>%
  group_by(dataset) %>%
  summarize(across(starts_with("diff_"), mean))

max.cent.diffs <- cent_diffs %>%
  group_by(dataset) %>%
  summarize(across(starts_with("diff_"), max))
```

Plot average differences in centiles
```{r}
mean.diffs.long <- mean.cent.diffs %>%
  pivot_longer(cols=starts_with("diff_"), values_to="centile_diff", names_to = "pheno", names_prefix = "diff_") %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)))

ggplot(mean.diffs.long) +
  geom_density(aes(x=centile_diff, fill=dataset), alpha=0.5) +
  facet_grid(dataset~pheno_cat)

mean.diffs.long %>%
  dplyr::filter(dataset== "cf.gam" | dataset == "cf.gam_refA" | dataset == "cf.gamlss" | dataset == "cf.gamlss_refA") %>%
  ggplot() +
  geom_density(aes(x=centile_diff, fill=dataset), alpha=0.5) +
  scale_x_continuous(labels = label_comma())

mean.diffs.long %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  mutate(centile_diff = centile_diff*100) %>% #for visualization purposes
  ggplot() +
  geom_histogram(aes(x=centile_diff, fill=dataset), alpha=0.5) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_x_continuous(labels = label_comma()) + facet_wrap(~dataset, scales = "free_y")
```

Look at max difference per phenotype
```{r}
max.diffs.long <- max.cent.diffs %>%
  pivot_longer(cols=starts_with("diff_"), values_to="max_centile_diff", names_to = "pheno", names_prefix = "diff_") %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)))

max.diffs.long %>%
  #dplyr::filter(dataset== "cf.lm_refA" | dataset == "cf.gam_refA" | dataset == "cf.gamlss" | dataset == "cf.gamlss_refA") %>%
  ggplot() +
  geom_density(aes(x=max_centile_diff, fill=dataset), alpha=0.5) +
  scale_x_continuous(labels = label_comma()) + facet_wrap(~pheno_cat)

max.diffs.long %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  ggplot() +
  geom_density(aes(x=max_centile_diff, fill=dataset), alpha=0.5) +
  scale_x_continuous(labels = label_comma()) + facet_wrap(~dataset, scales = "free_y")
```

Test for differences in distribution
```{r}
summary(aov(centile_diff ~ dataset, data=mean.diffs.long))

attach(mean.diffs.long)
pairwise.t.test(x=centile_diff, g=dataset, p.adj = "bonf")
detach()
```

Seems like gamlss methods are significantly better than the other combat methods and not significantly different from raw centile scores (bonferroni-adjusted)

@@

```{r}
mean.cent.diffs.by.sex <- cent_diffs %>%
  group_by(dataset, sex) %>%
  summarize(across(starts_with("diff_"), mean)) %>%
  ungroup() %>%
  pivot_longer(cols=starts_with("diff_"), values_to="centile_diff", names_to = "pheno", names_prefix = "diff_") %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)))

@@

max.cent.diffs.by.sex <- cent_diffs %>%
  group_by(dataset, sex) %>%
  summarize(across(starts_with("diff_"), max)) %>%
  ungroup() %>%
  pivot_longer(cols=starts_with("diff_"), values_to="max_centile_diff", names_to = "pheno", names_prefix = "diff_") %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)))

@@

Other methods are fairly consistently underestimating male centiles and overestimating female centiles.

>>>>

```{r}
drop1.df <- fread("ukb_basic/drop1_tests.csv", stringsAsFactors = TRUE) %>%
  dplyr::select(-V1)
sex.df <- fread("ukb_basic/sex_summary.csv", stringsAsFactors = TRUE) %>%
  dplyr::select(-V1)
sum.df <- fread("ukb_basic/gamlss_summary.csv", stringsAsFactors = TRUE) %>%
  dplyr::select(-V1)
```

@@

sigma.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & term =="sexMale") %>%
  group_by(dataset, pheno_cat) %>%
  ggplot() +
  geom_histogram(aes(x=var_est, fill = dataset), color="black", alpha=0.5, position = "identity") +
  theme_linedraw() +
  facet_wrap(~pheno_cat, scales= "free")
```

@@

```{r, fig.height=5, fig.widt=5}
sigma.df %>%
  dplyr::filter(pheno_cat == "Global Volume" & term =="sexMale") %>%
  mutate(var_est.sig = case_when(sig.sum_bf.corr == TRUE ~ var_est,
                                 sig.sum_bf.corr == FALSE ~ 0)) %>%
  #group_by(dataset) %>%
  ggplot() +
  geom_col(aes(x=dataset, y = var_est.sig, fill=dataset), position="dodge") +
  #scale_fill_gradient2() +
  facet_grid(~pheno) + 
  theme_bw(base_size=12) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  theme(legend.position="none") +
  #theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_rect(color= "gray", fill = NA)) +
  ggtitle("Beta weight for effect of male sex on global volumes' variance")

sigma.df.glob <- sigma.df %>%
  dplyr::filter(pheno_cat == "Global Volume" & term =="sexMale")
table(sigma.df.glob$dataset, sigma.df.glob$sig.sum_bf.corr)
```

@@

table(combined_preds$dataset)
```
calc how much centiles change
```{r}
cent_diffs <- combined_preds %>%
  mutate(dataset = factor(dataset, levels = unique(dataset), ordered = FALSE)) %>%
  mutate(dataset = relevel(dataset, ref= "ukb_CN")) %>%
  arrange(dataset) %>%
  group_by(participant) %>%
  mutate(across(all_of(pheno_list), ~ abs(. - first(.)), .names = "abs.diff_{.col}")) %>%
  ungroup()

diff_vol_glob <- sapply(vol_list_global, function(x) paste0("abs.diff_", x))

mean.cent.diffs <- cent_diffs %>%
  group_by(dataset) %>%
  summarize(across(starts_with("abs.diff_"), mean)) %>%
  ungroup()

max.cent.diffs <- cent_diffs %>%
  group_by(dataset) %>%
  summarize(across(starts_with("abs.diff_"), max)) %>%
  ungroup()
#names(max.cent.diffs)
```

Plot average differences in centiles
```{r}
mean.diffs.long <- mean.cent.diffs %>%
  pivot_longer(cols=starts_with("abs.diff_"), values_to="centile_abs.diff", names_to = "pheno", names_prefix = "abs.diff_") %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)))

mean.diffs.long %>%
  dplyr::filter(dataset != "ukb_CN") %>%
ggplot() +
  geom_density(aes(x=centile_abs.diff, fill=dataset), alpha=0.5) +
  facet_grid(dataset~pheno_cat)

mean.diffs.long %>%
  dplyr::filter(dataset== "cf.gam" | dataset == "cf.gam_refA" | dataset == "cf.gamlss" | dataset == "cf.gamlss_refA") %>%
  ggplot() +
  geom_density(aes(x=centile_abs.diff, fill=dataset), alpha=0.5) +
  scale_x_continuous(labels = label_comma())

mean.diffs.long %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  mutate(centile_abs.diff = centile_abs.diff*100) %>% #for visualization purposes
  ggplot() +
  geom_histogram(aes(x=centile_abs.diff, fill=dataset), color="black", binwidth=0.1, alpha=0.5, position="identity") +
  theme_linedraw() +
  #theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_x_continuous(labels = label_comma()) +
  ggtitle("Mean absolute differences in centile scores across ComBat \nprocessing pipelines for each phenotype")
```

Look at max difference per phenotype
```{r, fig.height=6, fig.width=7}
max.diffs.long <- max.cent.diffs %>%
  pivot_longer(cols=starts_with("abs.diff_"), values_to="max_centile_abs.diff", names_to = "pheno", names_prefix = "abs.diff_") %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)))

max.diffs.long %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  mutate(max_centile_abs.diff = max_centile_abs.diff*100) %>% #for visuals
  ggplot() +
  geom_histogram(aes(x=max_centile_abs.diff, fill=dataset), color="black", alpha=0.5, position="identity") +
  theme_linedraw(base_size = 12) +
  scale_x_continuous(labels = label_comma()) + facet_wrap(~pheno_cat) +
  ggtitle("Maximum absolute differences in centile scores across \nComBat processing pipelines for each phenotype")
```

```{r}
max.diffs.long %>%
  mutate(max_centile_abs.diff = max_centile_abs.diff*100) %>% #for visuals
  dplyr::filter(dataset != "ukb_CN") %>%
  ggplot() +
  geom_histogram(aes(x=max_centile_abs.diff, fill=dataset), color="black", alpha=0.5, position="identity") +
  theme_linedraw(base_size = 12) +
  scale_x_continuous(labels = label_comma())+
  ggtitle("Maximum absolute differences in centile scores across \nComBat processing pipelines for each phenotype")
```

Test for differences in distribution
```{r}
summary(aov(centile_abs.diff ~ dataset, data=mean.diffs.long))

attach(mean.diffs.long)
pairwise.t.test(x=centile_abs.diff, g=dataset, p.adj = "fdr")
detach()
```

Seems like gamlss methods are significantly better than the other combat methods and not significantly different from raw centile scores (BH corrected)

@@

```{r}
mean.cent.diffs.by.sex <- cent_diffs %>%
  group_by(dataset, sex) %>%
  summarize(across(starts_with("abs.diff_"), mean)) %>%
  ungroup() %>%
  pivot_longer(cols=starts_with("abs.diff_"), values_to="centile_diff", names_to = "pheno", names_prefix = "abs.diff_") %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)))

@@

max.cent.diffs.by.sex <- cent_diffs %>%
  group_by(dataset, sex) %>%
  summarize(across(starts_with("abs.diff_"), max)) %>%
  ungroup() %>%
  pivot_longer(cols=starts_with("abs.diff_"), values_to="max_centile_diff", names_to = "pheno", names_prefix = "abs.diff_") %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)))

@@

No shift in abs. differences, looking at overall diff.
```{r, fig.height=7, fig.width=8}
cent.diffs2 <- combined_preds %>%
  mutate(dataset = factor(dataset, levels = unique(dataset), ordered = FALSE)) %>%
  mutate(dataset = relevel(dataset, ref= "ukb_CN")) %>%
  arrange(dataset) %>%
  group_by(participant) %>%
  mutate(across(all_of(pheno_list), ~ . - first(.), .names = "diff_{.col}")) %>%
  ungroup()
```

```{r}
max.cent.diffs.by.sex2 <- cent.diffs2 %>%
  group_by(dataset, sex) %>%
  summarize_at(vars(starts_with("diff")), list(max_value = ~.[which.max(abs(.))])) %>%
  ungroup() %>%
  pivot_longer(cols=starts_with("diff_"), values_to="max_centile_diff", names_to = "pheno", names_prefix = "diff_") %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)))

max.cent.diffs.by.sex2 %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  mutate(max_centile_diff = max_centile_diff *100) %>%
ggplot() +
  geom_density(aes(x=max_centile_diff, fill=sex), alpha=0.5) +
  facet_wrap(~dataset, scales="free")
```
Other methods are fairly consistently underestimating male centiles and overestimating female centiles.

```{r}
mean.cent.diffs.by.sex2 <- cent.diffs2 %>%
  group_by(dataset, sex) %>%
  summarize(across(starts_with("diff_"), mean)) %>%
  ungroup() %>%
  pivot_longer(cols=starts_with("diff_"), values_to="centile_diff", names_to = "pheno", names_prefix = "diff_") %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)))

mean.cent.diffs.by.sex2 %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  mutate(centile_diff = centile_diff *100) %>%
ggplot() +
  geom_density(aes(x=centile_diff, fill=sex), alpha=0.5) +
  facet_wrap(~dataset, scales="free")
```
